The Evolution of Phenotypically Invariant Genetic Networks

### 2017-11-14 kalman_walk_paper.tex
changes to perturb Kalman

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
Each $F_2$ hybrid will be homozygous for one parental allele on average at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
but will have higher variance.
The variance of $F_2$s can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$s will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
In a model of $p$ traits in which the optimal set $\allS$ has dimension $q$,
% if each trait is controlled by a single locus, $\omega = XXX$,
using the polygenic model of appendix \ref{apx:seg_cov}, $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

\plr{note: should be a plot of the components of this against $T/N_e$ later}

@@

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a^2 T/N_e} \approx 1 - 2 \omega a T/N_e$.
Since in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N$) is proportional to $N_e$,
so this predicts the rate of decrease of hybrid fitness to be \emph{independent} of population size
for small populations.
\plr{look up what $\sigma_N^2$ is at mut-drift equilib}

% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
In the other direction, consider large metapopulations (or isolated ``species complexes'')
among which heritable variation is strongly constrained by selection
(i.e., there is substantial recombination load),
so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a^2 T/N_e} \approx 1 - \omega a T/N_e$,
and could be extremely rapid if $a$ is large.

but not likely to be other combinations; make plot of \eqref{eqn:fit2} against $T/N_e$ (or even $T$?) in these three situations.

\begin{figure}
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop if $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N = \sigma_S = \gamma$
\textbf{(B)} $\sigma_N = \sigma_S = 0.1 \gamma$
\textbf{(C)} $0.1 \sigma_N = \sigma_S = \gamma$.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

\subsection*{Parameter estimates for real systems}
%\js{I think we should move this section?}
%To translate these results into real predictions, we need to know
%the strength of stabilizing selection on the phenotype,
%and the amount (and structure) of heritable variation in the genotype.
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

%% amount of heritable variation
%We quantify (roughly) the amount of heritable variation
%by $\sigma^2$, the genetic variance present in a population in a typical entry of $\Sys$.
%The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
%per change in concentration of $j$.
%It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
%than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
%In the context of transcription factor networks this may be affected 
%not only by the binding strength of molecule $j$ to the promoter region of gene $i$
%but also the effects of other transcription factors (e.g., cooperativity)
%and local chromatin accessibility \citep{stefflova2013cooperativity}.
%For this reason, 
%the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
%typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
%and single variants may affect many entries of $\Sys$ simultaneously.
%% (Recall that although these are best modeled through nonlinear terms,
%% by linearizing we essentially consider first-order effects.)
%On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
%in the presence of substantial molecular noise \citep{canalization,buffering},
%suggesting that substantial variation in the micro-scale dynamics we consider here
%may be necessary to produce relevant phenotypic effects downstream.
%\plr{replace "micro-scale" with sthg else or discuss earlier}

%% empirical data on sigma
%The amount and structure of this standing variation is established over long time scales
%by many factors, including
%mutation-selection balance, 
%shifts in the phenotypic optimum,
%and/or spatial variation in the optimum \citep{hansen1996translating}.
%Quantitative genetics models of mutation-selection balance 
%predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
%but it is unclear how well these predictions match reality \citep{johnson_barton}
%and how much they are expected to change over time \citep{arnold_changing_G}.
%However, empirical work allows us to estimate at least the rough magnitude of variation.
%Differences in $A_{ij}$ due to a sequence change are hard to measure,
%but variation in both transcription factor binding site occupancy
%and expression levels (e.g., cis-eQTL) have been measured in various systems.
%However, variation in binding site occupancy may overestimate variation in $\Sys$,
%since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
%and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
%%Nonetheless,
%%\citet{kasowski2010variation} found differential occupancy in 7.5\%
%%of binding sites of a transcription factor (p65) between human individuals.
%%\citet{verlaan2009targeted} showed that cis-regulatory variation
%%accounts for around 2--6\% of expression variation in human blood-derived primary cells,
%%while \citep{lappalainen2013transcriptome} found that human population variation 
%%explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
%%Taken together, this suggests that variation in the entries of $\Sys$
%%may be on the order of 1\% between individuals of a population --
%%doubtless varying substantially between species and between genes.
%%
%Given the empirical data on transcription factor binding and gene expression
%intrapopulation variation \citet{kasowski2010variation, verlaan2009targetedm lappalainen2013transcriptome}, we conservatively estimate that variation in the regulatory coefficients in $\Sys$ may be on the order of 1\% between individuals of a population -- 
%doubtless varying substantially between species and between genes.
%
%%% argument for strength of selection
%%It seems certain that selection in most species is not so strong that intra-population variation
%%is strongly deleterious,
%%so that if $\beta$ is the typical scale on which selection acts,
%%then $\beta > \sigma$.  
%%However,
%%a range of studies \plr{find them} have found evidence for weak stabilizing selection
%%on regulatory SNPs and cis-eQTL.
%For instance, 
%\citet{maria_and_sergey} \plr{(others?)} found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
%This suggests that 
%the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
%so that perhaps $\sigma$ and $\beta$ are relatively close.
%This is as would be expected if available variation is held in check by mutation--selection balance
%rather than genetic drift.
%A conservative estimate would be that $\beta = 5 \sigma$;
%taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
%to effect a noticeable drop in fitness.
%\plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}
%
%We have guessed that within a population, 
%the entries of $\Sys$ vary by around 1\%, 
%at least for networks whose function is strongly constrained.

%%%%%%%%%%%%
\plri{insert commented-out discussion of what the actual genetic basis is}

@@

@@

@@

\jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}
\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.

>>>>

\paragraph{Hybridization}
The means of two allopatric populations each of effective size $N_e$ separated for $T$ generations
will be a distance roughly of order $2\sigma_N \sqrt{T/N_e}$ apart along $\optx$.
(Consult figure \ref{fig:conceptual_fig} for a conceptual diagram.)
A population of $F_1$ hybrids has one haploid genome from each,
whose coefficients are averaged,
and so will have mean system coefficients at the midpoint between their means.
Each $F_2$ hybrid will be homozygous for one parental allele on average at half of the loci in the genome,
so the distribution of $F_2$s will have mean at the average of the two populations,
but will have higher variance.
The variance of $F_2$s can be shown to increase linearly with the square of the distance between
parental population means
under models of both simple and polygenic traits.
This is suggested by figure \ref{fig:conceptual_fig} and shown in Appendix \ref{apx:seg_cov}.
\plr{connect to Barton etc polygenic adaptation lit}
Concretely, we expect the population of $F_1$s to have variance $\sigma^2_S$ in the selected direction
(the same as within each parental population),
but the population of $F_2$s will have variance of order $\sigma^2_S + 4 \omega \sigma^2_N T/N_e$,
where $\omega$ is a factor that depends on the genetic basis of the coefficients.
In a model of $p$ traits in which the optimal set $\allS$ has dimension $q$,
% if each trait is controlled by a single locus, $\omega = XXX$,
using the polygenic model of appendix \ref{apx:seg_cov}, $\omega = (p-q)/8$.
If each trait is controlled by a single locus, as in figure \ref{fig:conceptual_fig},
the value is similar.

@@

@@

%  $\sigma_N \approx \sigma_S \ll \gamma$, ``isolated population''
Consider instead a much smaller, isolated population
whose genetic variation is primarily constrained by genetic drift,
so that $\sigma_N \approx \sigma_S \ll \gamma$.
Setting $a = (\sigma_N/\gamma)^2$ to be small,
the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 4 \omega a T/N_e} \approx 1 - 2 \omega a T/N_e$.
Hybrid fitness seems to drop more slowly in this case in figure \ref{fig:speciation_rates},
but since time is scaled by $N_e$, so speciation may occur \emph{faster} than in a large population.
However, at least in some models \citep{bartonXXX}, in small populations at mutation-drift equilibrium
the amount of genetic variance ($\sigma_N^2$) is proportional to $N_e$,
which would compensate for this difference, 
perhaps even predicting the rate of decrease of hybrid fitness to be \emph{independent} of population size
for small populations.

% $\sigma_N \gg \sigma_S \approx \gamma$, if $N_e$ is big and there's a lot of recombination load (``species complex'')
In the other direction, consider large metapopulations (or isolated ``species complexes'')
among which heritable variation is strongly constrained by selection
(i.e., there is substantial recombination load),
so that $\sigma_S \approx \gamma$ but $\sigma_N/\gamma$ is large.
Then the fitness of $F_2$s is $\fit_2 \le 1/\sqrt{1 + 2 \omega a T/N_e} \approx 1 - \omega a T/N_e$,
and could be extremely rapid if $a$ is large.

For instance, in a population of one million organisms that has 10 generations per year (a drosophilid species, perhaps)
under the ``large population'' scenario of Figure \ref{fig:speciation_rates}A,
system drift would lead to a substantial fitness drop of around 10\% in $F_2$ hybrids in only 10,000 years.
This drop may be enough to induce evolutionary reinforcement of reproductive isolation.
If one thousand of these organisms is isolated (perhaps on an island, as in Figure~\ref{fig:speciation_rates}B),
then a similar drop could occur in around 120 years.
On the other hand, if the population is one of several of similar size
that have recently come into secondary contact after population re-expansion,
the situation may be similar to that of Figure~\ref{fig:speciation_rates}C with $N_e = 10^6$,
the same drop could occur after 1,100 years.
However, hyperdiverse populations of this type may not be stable on these time scales.

\begin{figure}[H]
\label{fig:speciation_rates}
\begin{center}
\includegraphics{speciation_rates}
\caption{
Mean drop if $F_2$ fitness relative to parental species,
with $\omega=1$ and 
\textbf{(A)} $\sigma_N^2 = \sigma_S^2 = \gamma^2$
\textbf{(B)} $\sigma_N^2 = \sigma_S^2 = 0.1 \gamma^2$
\textbf{(C)} $0.1 \sigma_N^2 = \sigma_S^2 = \gamma^2$.
The total fitness is from equation \eqref{eqn:fit2},
and is the product of the ``shift of mean'' and ``increase of variance'' terms
(the exponential and the square root, respectively).
Note that, as we assume, the contribution of the shift in mean
is small relative to that of increased variance.
\label{fig:speciation_rates}}
\end{center}
\end{figure}

\subsection*{Genetic variation in empirical regulatory systems}

Above we show how system variation and drift can contribute to gene network incompatibility.
What is known about the key quantity above, the amount of heritable variation in real regulatory networks?
%These are known at best only roughly \citep{felsenstein1988phylogenies},
%so we aim for order-of-magnitude estimates.

%% What is A and what determines its genetic variance.
The coefficient $A_{ij}$ from the system \eqref{eqn:system} measures how much the rate of net production of $i$ changes
per change in concentration of $j$.
It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
In the context of transcription factor networks this may be affected 
not only by the binding strength of molecule $j$ to the promoter region of gene $i$
but also the effects of other transcription factors (e.g., cooperativity)
and local chromatin accessibility \citep{stefflova2013cooperativity}.
For this reason, 
the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region,
and single variants may affect many entries of $\Sys$ simultaneously.
% (Recall that although these are best modeled through nonlinear terms,
% by linearizing we essentially consider first-order effects.)
On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
in the presence of substantial molecular noise \citep{canalization,buffering},
suggesting that substantial variation in the micro-scale dynamics we consider here
may be necessary to produce relevant phenotypic effects downstream.
\plr{replace "micro-scale" with sthg else or discuss earlier}

%% empirical data on sigma
The amount and structure of this standing variation is established over long time scales
by many factors, including
mutation-selection balance, 
shifts in the phenotypic optimum,
and/or spatial variation in the optimum \citep{hansen1996translating}.
Quantitative genetics models of mutation-selection balance 
predict precise levels and structure of standing variation \citep{kimura_mutsel,lande_mutsel,lande1981models},
but it is unclear how well these predictions match reality \citep{johnson_barton}
and how much they are expected to change over time \citep{arnold_changing_G}.
However, empirical work allows us to estimate at least the rough magnitude of variation.

Differences in $A_{ij}$ due to a sequence change are hard to measure,
but variation in both transcription factor binding site occupancy
and expression levels (e.g., cis-eQTL) have been measured in various systems.
However, variation in binding site occupancy may overestimate variation in $\Sys$,
since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin),
and variation in expression levels measures changes in steady-state concentration (our $\kappa_i$) rather than the \emph{rate} of change.
Nonetheless,
\citet{kasowski2010variation} found differential occupancy in 7.5\%
of binding sites of a transcription factor (p65) between human individuals.
\citet{verlaan2009targeted} showed that cis-regulatory variation
accounts for around 2--6\% of expression variation in human blood-derived primary cells,
while \citep{lappalainen2013transcriptome} found that human population variation 
explained about 3\% of expression variation.
%%\plr{Get some data from at least one other species in here!}
Taken together, this suggests that variation in the entries of $\Sys$
may be on the order of 1\% between individuals of a population --
doubtless varying substantially between species and between genes.

What matters is $\sigma$ relative to $\gamma$ but this is harder to know.

@@

Do we talk about the $G$ matrix and mutational correlation somewhere?

@@

Note that islands or bottlenecks leave large $\sigma_N$ but small $N_e$ so may go fast.

@@

\jss{talk about speciation rate in small population results as compared to Katari and Goldstein (I think). they look at the influence of $N_e$ on speciation rate.}

This has implications for genetic load also!!

hominids; neanderthal load.

\section*{Acknowledgements}
We would like to thank Sergey Nuzhdin, Stevan Arnold, Erik Lundgren, and Hossein Asgharian for valuable discussion.
